---
title: "Farmington and Bear River Bay Al criteria recalculation"
author: "Jake Vander Laan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true  
    toc_float: true
    fig_width: 8
    fig_height: 8
---

# Introduction

This document demonstrates how the [EPA 2018 aluminum criterion](https://www.epa.gov/wqc/2018-final-aquatic-life-criteria-aluminum-freshwater){target="_blank"} varies across pH, hardness, and DOC. 
It also pilots the application of the [species deletion procedure](https://www.epa.gov/sites/production/files/2015-08/documents/revised_deletion_process_for_the_site-specific_recalculation_procedure_for_aquatic_life_criteria.pdf){target="_blank"} 
to this criterion for Farmington and Bear River Bays of Great Salt Lake, showing the potential difference 
between the original Al criteria and recalculated criteria based on Farmington Bay and Bear River Bay taxa. EPA published Al criterion calculator R code was wrapped into a function 
within the DWQs wqTools R package to facilitate the calculation of the criterion across large datasets.

# Libraries

```{r, libraries}
library(dplyr)
library(wqTools)
```

# Test data

Generating test data for pH (6 to 9.5), hardness (0 to 400 mg/L), and DOC (0 to 5 mg/L). This data allows application of the Al criterion calculator across a wide range of conditions to see how each factor impacts the criteria and how 
the impact of the species deletion procedure and recalculation across a wide range of conditions.  

```{r, test-data}
pH=seq(6, 9.5, 0.5)
Hardness_mgL=seq(0, 400, 50)
DOC_mgL=seq(0, 5, 0.5)
test_data=merge(pH, Hardness_mgL)
names(test_data)=c('pH', 'Hardness_mgL')
test_data=merge(test_data, DOC_mgL)
names(test_data)=c('pH', 'Hardness_mgL', 'DOC_mgL')
```

# Unmodified criteria

```{r, calc-orig}
orig_crit=calcAlCrit(test_data, 'pH', 'Hardness_mgL', 'DOC_mgL')
```

```{r, echo=F, cmc-cfs, fig.cap="Figure 1. Unmodified CMC variability on correction factors."}
par(mfrow=c(2,2))
plot(Final_CMC~pH, orig_crit, ylab="CMC ug/L", xlab="pH")
plot(Final_CMC~Hardness_mgL, orig_crit, ylab="CMC ug/L", xlab="Hardness (mg/L)")
plot(Final_CMC~DOC_mgL, orig_crit, ylab="CMC ug/L", xlab="DOC (mg/L)")
par(mfrow=c(1,1))
```

```{r, echo=F, ccc-cfs, fig.cap="Figure 2. Unmodified CCC variability on correction factors."}
par(mfrow=c(2,2))
plot(Final_CCC~pH, orig_crit, ylab="CCC ug/L", xlab="pH")
plot(Final_CCC~Hardness_mgL, orig_crit, ylab="CCC ug/L", xlab="Hardness (mg/L)")
plot(Final_CCC~DOC_mgL, orig_crit, ylab="CCC ug/L", xlab="DOC (mg/L)")
par(mfrow=c(1,1))
```



# Species deletion & recalculation

Species deleted:  
1. Fatmucket, Lampsilis siliquoidea  
2. Snail, Melanoides tuberculata  


## Modify tox table

```{r, mod-tt}
tox_table_mod=dget('dput_tox_tables.txt')
tox_table_mod=subset(tox_table_mod, !Species %in% c('Fatmucket, Lampsilis siliquoidea', 'Snail, Melanoides tuberculata'))
```

## Calculate modified criteria

```{r, calc-mod}
mod_crit=calcAlCrit(test_data, 'pH', 'Hardness_mgL', 'DOC_mgL', tox_table=tox_table_mod)
```

```{r, echo=F, mod-cmc, fig.cap="Figure 3. Species deletion modified versus unmodified Al CMC. Dashed line = 1:1 line. Solid line = regression line.", fig.width=6, fig.height=6}
plot(mod_crit$Final_CMC~orig_crit$Final_CMC, main='CMC', xlab='Unmodified Al CMC', ylab='Recalculated Al CMC')
abline(0,1, lwd=2, lty=2)
lm=summary(lm(mod_crit$Final_CMC~orig_crit$Final_CMC))
abline(lm$coefficients[[1]], lm$coefficients[[2]], lwd=2)
```

```{r, echo=F, mod-ccc, fig.cap="Figure 4. Species deletion modified versus unmodified Al CCC. Dashed line = 1:1 line. Solid line = regression line.", fig.width=6, fig.height=6}
plot(mod_crit$Final_CCC~orig_crit$Final_CCC, main='CCC', xlab='Unmodified Al CCC', ylab='Recalculated Al CCC')
abline(0,1, lwd=2, lty=2)
lm=summary(lm(mod_crit$Final_CCC~orig_crit$Final_CCC))
abline(lm$coefficients[[1]], lm$coefficients[[2]], lwd=2)
```

# Discussion points

1. Deleted species for the FB & BRB recalculated Al criteria are not particularly sensitive taxa.  
2. Recalculation for FB and BRB species has a relatively small impact on Al criteria.  
3. Data points and the regression lines in figures 3 & 4 primarily fall below the 1:1 line, showing that recalculated criteria tend to be lower than the unmodified criteria under almost all combinations of correction factors.








